---
title: "R Flight Data EDA"
author: "Ewan Mitchell"
output:
  html_document:
    df_print: paged
fig_width: 15
fig_height: 15
---

```{r}
library(readr)
library(dplyr)       # For data manipulation
library(ggplot2)     # For data visualization
library(summarytools) # For summary statistics
library(ggplot2)
library(tidyr)
library(knitr)
library(maps)
library(mapdata)
library(corrplot)
library(factoextra)
library(RColorBrewer)
```

## 2019 Airline Delays with Weather and Airport Detail
This data set is about flight delays in the United States, providing  detailed information about airline, weather, airport and employment conditions. The aim of exploring this data set is to study the correlation between features and flight delays with the eventual goal, in Assessment 1, of performing a binary classification on this data set.


Data set available at "https://www.kaggle.com/datasets/threnjen/2019-airline-delays-and-cancellations".

```{r}
flightData <- read_csv("C:/Users/USER/OneDrive/Bristol Year 4/Data Science Toolbox/Group Assessments/full_data_flightdelay.csv")
```

We present the features of the data set. 
```{r}
colnames(flightData) 
```


```{r, results='asis'}

column_descriptions <- data.frame(
  Variable = c(
    "MONTH", "DAY_OF_WEEK", "DEP_DEL15", "DEP_TIME_BLK", "DISTANCE_GROUP", 
    "SEGMENT_NUMBER", "CONCURRENT_FLIGHTS", "NUMBER_OF_SEATS", "CARRIER_NAME", 
    "AIRPORT_FLIGHTS_MONTH", "AIRLINE_FLIGHTS_MONTH", "AIRLINE_AIRPORT_FLIGHTS_MONTH", 
    "AVG_MONTHLY_PASS_AIRPORT", "AVG_MONTHLY_PASS_AIRLINE", "FLT_ATTENDANTS_PER_PASS", 
    "GROUND_SERV_PER_PASS", "PLANE_AGE", "DEPARTING_AIRPORT", "LATITUDE", 
    "LONGITUDE", "PREVIOUS_AIRPORT", "PRCP", "SNOW", "SNWD", "TMAX", "AWND"
  ),
  Description = c(
    "Month", "Day of Week", "(TARGET) Departure Delay Indicator, 15 Minutes or More (1=Yes)", 
    "CRS Departure Time Block, Hourly Intervals", "Distance group to be flown by departing aircraft", 
    "The segment that this tail number is on for the day", "Concurrent flights leaving from the airport in the same departure block", 
    "Number of seats on the aircraft", "Carrier", "Avg Airport Flights per Month", 
    "Avg Airline Flights per Month", "Avg Flights per month for Airline AND Airport", 
    "Avg Passengers for the departing airport for the month", "Avg Passengers for airline for month", 
    "Flight attendants per passenger for airline", "Ground service employees (service desk) per passenger for airline", 
    "Age of departing aircraft", "Departing Airport", "Latitude of departing airport", 
    "Longitude of departing airport", "Previous airport that aircraft departed from", 
    "Inches of precipitation for day", "Inches of snowfall for day", 
    "Inches of snow on ground for day", "Max temperature for day", "Max wind speed for day"
  )
)

kable(column_descriptions, format = "html", caption = "Column Descriptions")
```

```{r}
glimpse(flightData)

head(flightData)

summary(flightData)

missing_values <- flightData %>%
  summarise_all(~ sum(is.na(.)))

missing_values
```

### Examining the proportion of delays by carrier.
```{r}
# Bar chart of carrier vs proportion of delays
delayed_proportions <- flightData %>%
  group_by(CARRIER_NAME) %>%
  summarize(proportion_delayed = mean(DEP_DEL15 == 1))

ggplot(delayed_proportions, aes(x = CARRIER_NAME, y = proportion_delayed)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    title = "Proportion of Delayed Flights by Carrier",
    x = "Carrier",
    y = "Proportion Delayed"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

We see that "Frontier Airlines" and "JetBlue Airways" have the highest proportions of delayed flights, while "Hawaiian Airlines" has the lowest.

### Examining the proportion of delays by month of the year.
```{r}

delayed_proportions <- flightData %>%
  group_by(MONTH) %>%
  summarize(proportion_delayed = mean(DEP_DEL15 == 1))

#flightData$MONTH <- factor(flightData$MONTH, levels = month.abb)

ggplot(delayed_proportions, aes(x = MONTH, y = proportion_delayed)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    title = "Proportion of Delayed Flights by Month",
    x = "Month",
    y = "Proportion Delayed"
  ) +
  theme_minimal() + 
 scale_x_discrete(limits = month.abb)

```

We can observe a small peak at June, which could be due to the start of the holiday season., while September and November have the lowest - which are typically not holiday months.


### Examining the proportion of delays by day of the week.
```{r}
delayed_proportions <- flightData %>%
  group_by(DAY_OF_WEEK) %>%
  summarize(proportion_delayed = mean(DEP_DEL15 == 1))

day_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

#flightData$DAY_OF_WEEK <- factor(flightData$DAY_OF_WEEK, levels = day_order)

ggplot(delayed_proportions, aes(x = DAY_OF_WEEK, y = proportion_delayed)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    title = "Proportion of Delayed Flights by Day of the Week",
    x = "Day of the Week",
    y = "Proportion Delayed"
  ) +
  theme_minimal() + 
  scale_x_discrete(limits = day_order)

```

The appears to only be slight variation between days of the week.

### Examining the number of flights by departure time block.
```{r}
# Histogram of DEP_TIME_BLK
ggplot(flightData, aes(x = DEP_TIME_BLK)) +
  geom_bar(fill = "blue") +
  labs(
    title = "Distribution of Flights by Departure Time Block",
    x = "Derpature Time Block",
    y = "Count"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

We see that there is a slight peak of flights at departure blocks between 6am - 9am, while between 9am -9pm it appears to be roughly uniform and with very few delays between 9pm - 6am. 

### Examining the proportion of delays by departure block.
```{r}

delayed_flights <- flightData %>%
  filter(DEP_DEL15 == 1)

# Create a histogram of DEP_TIME_BLK for delayed flights
ggplot(delayed_flights, aes(x = DEP_TIME_BLK, fill = DEP_TIME_BLK)) +
  geom_bar() +
  labs(
    title = "Distribution of Delayed Flights by Departure Time Block",
    x = "Departure Time Block",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We observe that this distribution looks like a skewed Gaussian distribution about a peak at 5pm-6pm which experiences the highest proportion of delayed flights.

### Density plot of Departing Airplorts with Delayed flights
```{r}

us_map <- map_data("usa") # map of the US

map_plot <- ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), fill = NA, color = "black", alpha = 0.5) +
  coord_fixed(ratio = 1)   # Ensure aspect ratio is 1:1

ggplot(delayed_flights, aes(x = LONGITUDE, y = LATITUDE)) +
  geom_density_2d() +
  coord_fixed(ratio = 1) +  # Ensure aspect ratio is 1:1
  labs(title = "Density Plot of Departing Airports with Delayed Flights") +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), fill = "white", color = "black", alpha = 0.5) 
# (slow)

```

We see that there are hot spots particularly in the east coast and west coast experiencing the highets density of flight delays.

### Examining the proportion of delayed flights by departing airport
```{r, fig.width=17}

# Calculate the count of delayed flights by departing airport
delayed_counts <- delayed_flights %>%
  group_by(DEPARTING_AIRPORT) %>%
  summarize(delayed_count = n())

total_counts <- flightData %>%
  group_by(DEPARTING_AIRPORT) %>%
  summarize(total_count = n())

airport_proportions <- left_join(delayed_counts, total_counts, by = "DEPARTING_AIRPORT") %>%
  mutate(proportion_delayed = delayed_count / total_count) %>%
  arrange(desc(proportion_delayed))

ggplot(airport_proportions, aes(x = reorder(DEPARTING_AIRPORT, -proportion_delayed), y = proportion_delayed)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    title = "Proportion of Delayed Flights by Departing Airport",
    x = "Departing Airport",
    y = "Proportion Delayed"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

We now use two methods to study the upper outliers of this graph.

```{r}

 # IQR method
iqr <- IQR(airport_proportions$delayed_count)
q3 <- quantile(airport_proportions$delayed_count, 0.75)

upper_bound <- q3 + 1.5 * iqr

outliers <- airport_proportions %>%
  filter(delayed_count > upper_bound)

print(outliers)

# z-score method

outliers <- airport_proportions %>%
  filter(scale(delayed_count) > 1.96) 

print(outliers)
```

### Examining the correlation between airport information + flight delays
We look at only numeric features that concern the departing airport and airline and the flight departures, and perform correlation analysis and PCA.

```{r}

airport_vars <- c(
  "AIRPORT_FLIGHTS_MONTH", "AIRLINE_FLIGHTS_MONTH", "AIRLINE_AIRPORT_FLIGHTS_MONTH",
  "AVG_MONTHLY_PASS_AIRPORT", "AVG_MONTHLY_PASS_AIRLINE",
  "FLT_ATTENDANTS_PER_PASS", "GROUND_SERV_PER_PASS", "PLANE_AGE",
  "LATITUDE", "LONGITUDE", "DEP_DEL15"
)

filtered_data <- flightData %>%
  select(all_of(airport_vars))

corr <- cor(filtered_data)

corrplot(corr, method = "color", type = "upper", order = "hclust", tl.cex = 0.6, title = "Correlation Plot of airport Features + DEP_DEL15")


```

```{r}

outpca <- prcomp(filtered_data, center = TRUE, scale. = TRUE)

p1 <- fviz_eig(outpca)
p1

p2 <- fviz_pca_var(outpca,
                   axes = c(1, 2), 
                   col.var = "contrib", 
                   gradient.cols = brewer.pal(n = 10, name = "RdBu"),
                   repel = TRUE
)
p2

```

We do not observe any strong correlation between the flight delays variable - "DEP_DEL15" and the other variables.

### Examining the correlation between weather information + flight delays
We look at only numeric features that concern the weather and the flight departures, and perform correlation analysis and PCA.

```{r}
weather_vars <- c("PRCP",	"SNOW",	"SNWD",	"TMAX", "AWND", "DEP_DEL15")

filtered_data <- flightData %>%
  select(all_of(weather_vars))

corr <- cor(filtered_data)

corrplot(corr, method = "color", type = "upper", order = "hclust", tl.cex = 0.6, title = "Correlation Plot of Weather Features + DEP_DEL15")
```

```{r}

outpca <- prcomp(filtered_data, center = TRUE, scale. = TRUE)

p1 <- fviz_eig(outpca)
p1

p2 <- fviz_pca_var(outpca,
                   axes = c(1, 2), 
                   col.var = "contrib", 
                   gradient.cols = brewer.pal(n = 10, name = "RdBu"),
                   repel = TRUE
)
p2

```

For instance, we can clearly observe a positive correlation between "PRCP" - Inches of precipitation for day and flight delays and also a slightly smaller psoitive correlation between "AWND" - max wind speed for day and flight delays.
